### 2015-06-19 chapters/ipa_meets_unicode.tex
refinements of IPA chapter

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode standard, which is probably best seen as an error.

\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.

\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.

\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive ``sucking'' sound similar
      to a fricative, for example <kǃ͓ʰ>; second, ``frictionalized'' vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.

\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
      \textbf{WHAT EXACTLY IS THIS?}

\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.

\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ``ejective'' or ``creaky voicing'' are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.

\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.

\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.

@@

Summarizing the pitfalls discussed in this chapter, we propose to define a 
``strict'' IPA encoding to be used when interoperability of phonetic resources 
is intended. For regular research papers, when only the visual outline is 
important,

\input{tables/unicode_letters}

\input{tables/unicode_diacritics}

\input{tables/unicode_boundaries}

\input{tables/unicode_additions}

>>>>

\item \textsc{Retroflex click} \newline
      Retroflex clicks can be represented by <‼> \textsc{double exclamation
      mark} at \uni{203C}. Note that the (post-)alveolar click <ǃ> at \uni{01C3}
      is confusingly referred to as \textsc{latin letter retroflex click} in the
      Unicode standard, which is probably best seen as an error.
\item \textsc{Voiced retroflex implosive} \newline 
      Although the IPA includes a
      series of voiced implosives (marked with a hook on top, see
      Section~\ref{pitfall-missing-decomposition}), there is no voiced retroflex
      implosive. Following the spirit of the IPA, we propose to use <\charis{ᶑ}>
      \textsc{latin small letter d with hook and tail} at \uni{1D91} for this
      sound.
\item \textsc{Fortis/lenis} \newline
      Languages described as having a fortis/plain/lenis distinction that
      corresponds poorly with the traditional
      voiced/voiceless-unaspirated/voiceless-aspirated continuum can be marked
      using the voiceless glyph for the plain phoneme, and then 
      <\dia{0348}> \textsc{combining double vertical line below} at
      \uni{0348} to mark the fortis articulation, and/or <\dia{0349}>
      \textsc{combining left angle below} at \uni{0349} for the lenis
      articulation.
\item \textsc{Frictionalization} \newline 
      The diacritic <\dia{0353}>
      \textsc{combining x below} at \uni{0353} can be used to represent three
      types of frictionalized sounds: First, click consonants where the release
      of the anterior closure involves an ingressive ``sucking'' sound similar
      to a fricative, for example <kǃ͓ʰ>; second, ``frictionalized'' vowels
      (sounds that are phonologically vocalic, but with sufficiently close
      closures to create buzzing); and third, fricative sounds at places of
      articulation that do not have dedicated fricative glyphs, for example
      sounds with voiceless velar lateral frication, like <ʟ̥͓>.
\item \textsc{Derhoticization} \newline 
      For derhoticization we propose to use
      <\dia{032E}> \textsc{combining breve below} at \uni{032E}. 
      \textbf{WHAT EXACTLY IS THIS?}
\item \textsc{Coronal non-sibilant} \newline
      Languages described as having a sibilant/non-sibilant distinction among
      coronal fricatives and affricates can be handled using the subscript
      <\dia{0347}> \textsc{combining equals sign below} at \uni{0347} to mark
      the non-sibilant phoneme.
\item \textsc{Glottalization} \newline 
      Glottalized sounds can be indicated using
      <\dia{02C0}> \textsc{modifier letter glottal stop} at \uni{02C0}, unless
      it is clear that either ``ejective'' or ``creaky voicing'' are the
      intended sounds (in which cases the standard IPA diacritics should be
      used). Pre-glottalized sounds can be marked with
      <\diareverse{02C0}> to the left of the base
      glyph, for example <\charis{ˀt}>.
\item \textsc{Voiced pre-aspiration} \newline Voiced sounds having
      pre-aspiration can be marked with
      <\diareverse{02B1}> \textsc{modifier letter
      small h with hook} at \uni{02B1} to the left of the base glyph, for
      example <\charis{ʱd}>.
\item \textsc{Epilaryngeal phonation} \newline 
      There are some rare articulations that make
      use of an epilaryngeal phonation mechanism (e.g.,~the “sphincteric vowels”
      of~!Xóõ). To represent these vowels, we propose to use the modifier <\dia{1D31}>
      \textsc{modifier letter capital e} at \uni{1D31} to denote such sphincteric
      phonation.

@@

Summarizing the pitfalls as discussed in this chapter, we propose to define a 
``strict'' IPA encoding to be used when interoperability of phonetic resources 
is intended. For regular research papers, when only the visual outline is 
important, the details encoding of IPA in electronic format is rather 
unimportant, as long as it looks as expected. So for those regular usage these 
proposals are irrelevant.

[TO DO: different kind of proposals]

At the end of this chapter we have added a few longish tables summarizing all
155 different unicode codepoints that form the basis of strict-IPA encoding
(107 letters, 41 diacritics and 7 remaining symbols). Each of
these tables shows a typical glyph, and then lists the Unicode Codepoint,
Unicode Name and IPA description for each symbol.

TO DO:
[ordering]
[removed characters: tie bar, tilde overlay]
[additions]

\begin{itemize}[itemsep=6pt]

  \item \textsc{IPA letters} \newline
        The 107 different IPA letters as allowed in strict-IPA encoding are
        listed in Table~\ref{tab:ipa_letters} starting on
        page~\pageref{tab:ipa_letters}.
  \item \textsc{IPA diacritics} \newline The 41 different IPA diacritics and
        tone markers (both Unicode Modifier Letters and Combining Diacritical
        Marks) as allowed in strict-IPA encoding are listed in
        Table~\ref{tab:ipa_diacritics} starting on
        page~\pageref{tab:ipa_diacritics}.
  \item \textsc{IPA remainders} \newline
        The 7 remaining IPA symbols (boundary and stress markers) as allowed in
        strict-IPA encoding are listed in Table~\ref{tab:ipa_boundaries} on
        page~\pageref{tab:ipa_boundaries}. 
  \item \textsc{IPA additions} \newline
        The 10 proposed additions to the IPA are listed in
        Table~\ref{tab:ipa_additions} on page~\pageref{tab:ipa_additions}.

\newpage
\input{tables/unicode_letters}
\input{tables/unicode_diacritics} 
\input{tables/unicode_boundaries}
\input{tables/unicode_additions}

### 2015-06-19 tables/unicode_boundaries.tex
refinements of IPA chapter

\begin{center}
\begin{xtabular}{ l L{1.2cm} L{5.6cm} L{4cm} }
{\large ˈ} & \unif{02C8} & \textsc{modifier letter vertical line} & primary stress \\
{\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
{\large.} & \unif{002E} & \textsc{full stop} & syllable break \\
{|} & \unif{007C} & \textsc{vertical line} & minor group break (foot) \\
{‖} & \unif{2016} & \textsc{double vertical line} & major group break (intonation) \\
\charis{‿} & \unif{203F} & \textsc{undertie} & linking (absence of a break) \\
\end{xtabular}
\end{center}

>>>>

\begin{center}
\begin{xtabular}{ l L{1.2cm} L{5.6cm} L{4cm} }
{\large ˈ} & \unif{02C8} & \textsc{modifier letter vertical line} & primary stress \\
{\large ˌ} & \unif{02CC} & \textsc{modifier letter low vertical line} & secondary stress \\
 & \unif{0020} & \textsc{space} & word break \\
{\large.} & \unif{002E} & \textsc{full stop} & syllable break \\
{|} & \unif{007C} & \textsc{vertical line} & minor group break (foot) \\
{‖} & \unif{2016} & \textsc{double vertical line} & major group break (intonation) \\
\charis{‿} & \unif{203F} & \textsc{undertie} & linking (absence of a break) \\
\end{xtabular}
\end{center}
